---
title: "Documentation and tutorials for *DHARMa diagnostics in closed-population capture–recapture
  models*"
author: "Documentation and tutorial supplement"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
  pdf_document:
    toc: true
    toc_depth: '3'
params:
  full_grid: true
  run_applications: true
  make_figures: true
  output_dir: outputs
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6,
  cache   = FALSE
)
set.seed(params$seed)
if (!dir.exists(params$output_dir))
  dir.create(params$output_dir, recursive = TRUE)
```

# Required packages and helper functions

```{r packages}
core_pkgs <- c("DHARMa", "VGAM", "mgcv", "secr", "dplyr", "tibble")
missing_pkgs <- core_pkgs[!vapply(core_pkgs, requireNamespace,
                                  logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0)
  stop("Install required packages: ", paste(missing_pkgs, collapse = ", "))

library(DHARMa)
library(VGAM)
library(mgcv)
library(secr)
library(dplyr)
library(tibble)

source("CR_GOF_Functions.R")
```

---

# Data-preparation functions (Web Appendix B)

`prepare_recap_data_Mh()` returns one row per observed individual, with
`Y_recap = Y_i - 1` (recaptures) and `n_opp = tau - t_i` (opportunities).

`prepare_recap_data_Mtbh()` returns one Bernoulli row per post-first-capture
occasion, with `cap_prev` (previous-occasion indicator) and `occasion` (factor).

```{r show-prep-functions, eval=FALSE}
# Binomial partial-likelihood data (M_h models).

pl_mh <- prepare_recap_data_Mh(cap_hist, covariates, tau)

# Bernoulli partial-likelihood data (M_t / M_b / M_bh / M_th / M_tbh).

pl_mtbh <- prepare_recap_data_Mtbh(cap_hist, covariates, tau)
```


# Real-data applications (Web Appendix C)

These chunks reproduce Examples 1 and 3 of Web Appendix C. All file paths are 
relative; set `run_applications: true` in the YAML parameters to execute them.

## Mountain Pygmy PossumData (Web Appendix C1)

Expected file: `poss2003.RData`, containing `data$x.obs` and `data$cap.hist`.

```{r application-possum, eval=params$run_applications}
if (!file.exists("poss2003.RData")) {
  message("Skipping possum example: poss2003.RData not found.")
} else {
  load("poss2003.RData")
  body.weight <- scale(data$x.obs[-c(2, 27)])
colnames(body.weight) <- "body.weight"

cap.hist <- data$cap.hist[-c(2, 27), ]

tau <- ncol(cap.hist)    # Number of capture occasions.

D <- nrow(cap.hist)      # Number of unique individuals captured.

#...............................................................................

# Test data for closure using the Stanley--Burnham test via the secr R-package.

captXY <- c()

for(i in 1: nrow(cap.hist)) {
  Occasion <- which (cap.hist[i, ] > 0)
  yy <- length(Occasion)
  ID <- rep(i, yy)
  Session <- rep(1, yy)
  ind1 <- cbind(Session, ID, Occasion)
  captXY <- rbind(captXY, ind1)
}

closure.test(make.capthist(captXY, traps = NULL, 
                           fmt = c("trapID"), noccasions = tau))

# A low p-value (typically less than 0.05) suggests that the population is NOT 
# closed and that an open population model (like the Jolly-Seber, CJS, POPAN 
# model) should be used instead.

# Here the P-value is 0.53320675, it's likely to be a closed population.

#...............................................................................

# Construct (Binomial) recapture-based quantities under the partial likelihood 
# for fitting M_h-only models.

pl_data <- prepare_recap_data_Mh(cap.hist, body.weight, tau)

# Fit a M_h GLM.

mod_mh_glm <- glm(cbind(Y_recap, n_opp - Y_recap) ~ body.weight, 
                 family = binomial(link = "logit"), data = pl_data)

residuals_output1 <- simulateResiduals(mod_mh_glm, n = 2000)

# Fit a M_h GAM.

mod_mh_gam <- gam(cbind(Y_recap, n_opp - Y_recap) ~ s(body.weight, bs = "ps"), 
                family = binomial(link = "logit"), gamma = 1.4, data = pl_data)

# Construct (Bernoulli) recapture-based quantities under the partial likelihood
# for fitting M_t- and M_b-type models.

pl_data <- prepare_recap_data_Mtbh(cap.hist, body.weight, tau)

# Fit an M_bh GAM (accounts for behave effects and hete.).

mod_mbh_gam <- gam(Y_recap ~ s(body.weight, bs = "ps") + cap_prev, 
               family = binomial(link = "logit"), gamma = 1.4, data = pl_data)

# Fit a M_tbh-type GAM (accounts for time, behave effects and hete.).

mod_mtbh_gam <- gam(Y_recap ~ s(body.weight, bs = "ps") + cap_prev + occasion, 
                family = binomial(link = "logit"), gamma = 1.4, data = pl_data)

model_names <- c("(a) Model: M_h GLM", "(b) Model: M_h GAM", 
                 "(c) Model: M_bh GAM", "(d) Model: M_tbh GAM")

models <- list(mod_mh_glm, mod_mh_gam, mod_mbh_gam, mod_mtbh_gam)

for(i in 1:4) {
  if(i == 1 | i == 2) {
    plot(DHARMa::simulateResiduals(models[[i]], n = 2000,
                                      main = paste("Model:", model_names[i])))
  }
  if(i == 3 | i == 4) {
    
    group_weight <- aggregate(body.weight ~ ID, data = pl_data, 
                              FUN = mean)$body.weight
    par(mfrow = c(1, 2))
    res_plot0 <- DHARMa::simulateResiduals(models[[i]], n = 2000,
                                            main = paste("Model:", 
                                                         model_names[i]))
    res_plot <- recalculateResiduals(res_plot0, group = pl_data$ID)
    
    plotQQunif(res_plot)
    plotResiduals(res_plot, group_weight)
  }
  
  mtext(paste("", model_names[i]), side = 3, line = -1.5, 
        outer = TRUE, adj = 0.1, cex = 1.7)
}

#...............................................................................

# Get the population size from fitting model a M_bh GAM using the partial 
# likelihood and the Huggins' conditional likelihood (CL).

mbh_study_grid <- expand.grid(ID = unique(pl_data$ID), occ = 1:tau)

# For Huggins M_bh, we usually assume cap_prev = 0 for the "potential" 
# probability of the first capture.

mbh_study_grid$body.weight <- body.weight[mbh_study_grid$ID] 
mbh_study_grid$cap_prev <- 0 

mbh_preds <- predict(mod_mbh_gam, newdata = mbh_study_grid, type = "response")
mbh_preds_list <- split(mbh_preds, mbh_study_grid$ID)

pi_i <- sapply(mbh_preds_list, function(x) 1 - prod(1 - x))
pi_i2 <- sapply(mbh_preds_list, function(x) prod(1 - x))

N_hat1 <- sum(1/pi_i)         # The Horvitz-Thompson estimator.

# To obtain the standard error for the N_hat1, we use a Taylor's series approx.
# see Stoklosa et al. (2011) for further details.

var.beta <- mod_mbh_gam$Ve
X_des <- model.matrix(mod_mbh_gam, newdata = mbh_study_grid)

pi02 <- (1 - pi_i)/pi_i/pi_i
cz0.dot <- absorb(X_des*c(mbh_preds), tau)
d1 <- apply(pi02*cz0.dot, 2, sum)
var_N_hat1 <- sum((1 - pi_i)/pi_i/pi_i) + d1%*%var.beta%*%d1

suppressWarnings(detach("package:mgcv", unload = TRUE))

library(VGAM)

data_Mbh_VGAM <- data.frame(cbind(cap.hist, body.weight))

colnames(data_Mbh_VGAM) <- c("y1", "y2", "y3", "y4", "y5", "body.weight")
  
M_bh_VGAM <- vgam(cbind(y1, y2, y3, y4, y5) ~ s(body.weight, df = 1.5), 
               posbernoulli.b, data = data_Mbh_VGAM, trace = FALSE) 

N_hat2 <- M_bh_VGAM@extra$N.hat

# Population size estimate (standard error) for M_bh using partial likelihood:

print(c(N_hat1, sqrt(var_N_hat1)))

# Population size estimate (standard error) for M_bh using conditional likelihood:

print(c(N_hat2, M_bh_VGAM@extra$SE.N.hat)) # Standard errors for N_hat.
}
```


## Edinburgh Taxicab Data (Web Appendix C3)

Expected file: `TaxiM.inp` (available from the MARK software repository).

Population size is estimated under $M_t$ (via VGAM conditional likelihood)
and $M_{th}$ (via `TMB_conditional_GLMM.cpp`, which implements the fully
marginalised conditional-likelihood GLMM described in Web Appendix D6).

```{r application-taxicab, eval=params$run_applications}
if (!file.exists("TaxiM.inp")) {
  message("Skipping taxicab example: TaxiM.inp not found.")
} else {
  
  cpp_file <- "TMB_eps_NEW2.cpp"
  dll_name <- "TMB_eps_NEW2"
  if (file.exists(cpp_file)) {
    TMB::compile(cpp_file, 
                 flags = "-Wno-ignored-attributes -O2 -mfpmath=sse -msse2 -mstackrealign")
    dyn.load(TMB::dynlib(dll_name))
    
    lines      <- readLines("TaxiM.inp")
    data_lines <- lines[grepl(";", lines)]
    ch_strings <- gsub(".*\\*/\\s*([01]+).*", "\\1", data_lines)
    cap.hist   <- matrix(as.numeric(unlist(strsplit(ch_strings, ""))),
                         ncol = nchar(ch_strings[1]), byrow = TRUE)
    tau <- ncol(cap.hist)   # tau = 10
    D   <- nrow(cap.hist)
    
    captXY <- c()
    
    for(i in 1: nrow(cap.hist)) {
      Occasion <- which (cap.hist[i, ] > 0)
      yy <- length(Occasion)
      ID <- rep(i, yy)
      Session <- rep(1, yy)
      ind1 <- cbind(Session, ID, Occasion)
      captXY <- rbind(captXY, ind1)
    }
    
    closure.test(make.capthist(captXY, traps = NULL, fmt = c("trapID"), 
                               noccasions = tau))
    
    # A low p-value (typically less than 0.05) suggests that the population is NOT 
    # closed and that an open population model should be used instead.
    
    # Here, the P-value is 0.4615, it's likely to be a closed population.
    
    #...............................................................................
    
    # Construct (Bernoulli) recapture-based quantities under the partial likelihood
    # for fitting M_t- and M_b-type models.
    
    pl_data1 <- prepare_recap_data_Mtbh(cap.hist, NULL, tau)
    
    # Fit a M_0 GLM (constant model).
    
    mod_m0_glm <- glm(Y_recap ~ 1, data = pl_data1, 
                      family = binomial(link = "logit"))
    
    # Fit a M_t GLM (accounts for time effects only).
    
    mod_mt_glm <- glm(Y_recap ~ occasion, data = pl_data1, 
                      family = binomial(link = "logit"))
    
    # Fit model M_t with VGAM (which uses the Huggins conditional likelihood) 
    # to get the population size estimate
    
    VGAM_data_Mt <- data.frame(cap.hist)
    colnames(VGAM_data_Mt) <- c("y1", "y2", "y3", "y4", "y5", 
                                "y6", "y7", "y8", "y9", "y10")
    
    M_t <- vglm(cbind(y1, y2, y3, y4, y5, y6, y7, y8, y9, y10) ~ 1, 
                posbernoulli.t, data = VGAM_data_Mt, trace = FALSE) 
    
    N_hat1 <- M_t@extra$N.hat
    
    model_names <- c("(a) Model: M_0 GLM", "(b) M_t GLM")
    models <- list(mod_m0_glm, mod_mt_glm)
    
    for(i in 1:2) {
      res_plot <- DHARMa::simulateResiduals(models[[i]], n = 1000,
                                            main = paste("Model:", model_names[i]))
      plot(recalculateResiduals(res_plot, group = pl_data1$ID))
      
      mtext(paste("", model_names[i]), side = 3, line = -2, 
            outer = TRUE, adj = 0.1, cex = 1.7)
    }
    
    #...............................................................................
    
    # Get the population size estimates from fitting model M_th (GLMM) using 
    # Huggins conditional likelihood. This model is a fully marginalized CL model
    # that uses Laplace approximation via TMB.
    
    tidbits_data <- list(y = as.matrix(cap.hist))
    
    tidbits_parameters <- list(
      time_effects = numeric(ncol(cap.hist)),
      individual_effects = rep(0, D),
      logsigma_individual_effects = log(1) 
    )
    
    tidbits_random <- c("individual_effects")
    
    objs <- TMB::MakeADFun(data = tidbits_data, 
                           parameters = tidbits_parameters, 
                           random = tidbits_random, 
                           DLL = dll_name, silent = TRUE)
    
    fit_RE_marg <- nlminb(start = objs$par, objective = objs$fn, gradient = objs$gr, 
                          control = list(trace = 0, iter.max = 5000, eval.max = 5000))
    
    fit_RE_marg_results <- TMB::sdreport(objs, bias.correct = FALSE, 
                                         ignore.parm.uncertainty = FALSE, 
                                         skip.delta.method = FALSE)
    fit_RE_marg_estimates <- as.list(fit_RE_marg_results, what = "Estimate", 
                                     report = TRUE)
    fit_RE_marg_all_results <- summary(fit_RE_marg_results, 
                                       select = "report", p.value = TRUE) %>% 
      as.data.frame() %>% tibble::rownames_to_column(var = "Parameters")
    
    # Get the population size estimate.
    
    N_hat2 <- fit_RE_marg_all_results %>% 
      dplyr::filter(Parameters == "population_size")
    
    # Population size estimate (standard error) for M_t:
    
    print(c(N_hat1, M_t@extra$SE.N.hat))
    
    # Population size estimate (standard error) for M_th:
    
    print(c(N_hat2$Estimate, N_hat2$`Std. Error`))
  }
}
```

